The zero--one laws of Kolmogorov and Hewitt--Savage in categorical probability

### 2019-10-29 categorical-zero-one.tex
Add hewitt-savage zero-one

A \emph{finite permutation} of the set $J$ is a bijection $\alpha : J \to J$ which fixes all but finitely many elements. Suppose that $\alpha : J \to J$ is a finite permutation or more generally any injection. Then an application of the universal property of the Kolmogorov power shows that $\alpha$ induces a morphism $\hat{\alpha} : X_J \to X_J$ by applying the universal property of $X_J$ to the compatible family of composite morphisms
\[
	\begin{tikzcd}
		X_J \ar{r}{\pi_F} & X_F \ar{r}{\cong} & X_{\alpha(F)},
	\end{tikzcd}
\]
where the second morphism matches up the factors of the two finite products involved as prescribed by the bijection $\alpha|_F : F \stackrel{\cong}{\longrightarrow} \alpha(F)$. The relevant compatibility condition is straightforward to check. Moreover, $\hat{\alpha}$ is deterministic by \Cref{kolmprod_catdet}.

@@

\begin{corollary}[Kolmogorov zero--one law]
	Suppose $\Omega$ is a standard Borel space with a probability measure $P$ and that $\left(f_j: \Omega \to X_j\right)_{j \in \bN}$ is a sequence of independent random variables taking values in standard Borel spaces $X_j$. Suppose that $T \subseteq \Omega$ is in the coarsest $\sigma$-algebra which makes the $f_j$ measurable, and independent of any finite subset of the $f_j$. Then $P(T) \in \{0,1\}$.
\end{corollary}
\begin{proof}
	We want to apply \Cref{thm:kolmog} with $\cC = \BorelStoch$ and $J = \bN$ and $A = I$, the one-element measurable space. Then $P : I \to \Omega$. We write $f_\bN : \Omega \to X_\bN$ for the induced map to the product. In $\Omega$, the coarsest $\sigma$-algebra which makes the $f_i$ measurable coincides with the $\sigma$-algebra consisting of the sets of the form $f_\bN^{-1}(S)$ for measurable $S \subseteq X_J$. Hence the assumption implies that there is measurable $S \subseteq X_J$ such that $T = f_\bN^{-1}(S)$. Then we apply \Cref{thm:kolmog} with $s : X_J \to \{0,1\}$ the indicator function of $S$, which is a measurable map and therefore can also be regarded as a morphism in $\BorelStoch$. We therefore obtain that $sp : I \to \{0,1\}$ is deterministic. But since $sp$ corresponds to the probability measure on $\{0,1\}$ with weight $P(f_\bN^{-1}(S)) = P(T)$ on $\{1\}$ and the complementary weight $1 - P(T)$ on $0$, we conclude that indeed $P(T) \in \{0,1\}$.
\end{proof}

@@

\todo[inline]{describe the application of \Cref{thm:hewsav} to $\BorelStoch$}

>>>>

A \emph{finite permutation} of the set $J$ is a bijection $\alpha : J \to J$ which fixes all but finitely many elements. Suppose that $\alpha : J \to J$ is a finite permutation or more generally any injection. Then an application of the universal property of the Kolmogorov power shows that $\alpha$ induces a morphism $\hat{\alpha} : X_J \to X_{\alpha(J)}$ by applying the universal property of $X_J$ to the compatible family of composite morphisms
\[
	\begin{tikzcd}
		X_J \ar{r}{\pi_F} & X_F \ar{r}{\cong} & X_{\alpha(F)},
	\end{tikzcd}
\]
where the second morphism matches up the factors of the two finite products involved as prescribed by the bijection $\alpha|_F : F \stackrel{\cong}{\longrightarrow} \alpha(F)$. Here we are using the fact that each finite subset of $\alpha(J)$ has the form $\alpha(F)$ for a unique finite $F \subset J$ (namely the preimage).
The relevant compatibility condition is straightforward to check. Moreover, $\hat{\alpha}$ is deterministic by \Cref{kolmprod_catdet}.

@@

\begin{corollary}[Kolmogorov zero--one law]
	Suppose $\Omega$ is a standard Borel space with a probability measure $P$ and that $\left(f_j: \Omega \to X_j\right)_{j \in \bN}$ is a sequence of independent random variables taking values in standard Borel spaces $X_j$. Suppose that $T \subseteq \Omega$ is in the coarsest $\sigma$-algebra which makes the $f_j$ measurable, and independent of any finite subset of the $f_j$. Then $P(T) \in \{0,1\}$.
\end{corollary}
\begin{proof}
	We want to apply \Cref{thm:kolmog} with $\cC = \BorelStoch$ and $J = \bN$ and $A = I$, the one-element measurable space. Then $P : I \to \Omega$. We write $f_\bN : \Omega \to X_\bN$ for the induced map to the product. In $\Omega$, the coarsest $\sigma$-algebra which makes the $f_i$ measurable coincides with the $\sigma$-algebra consisting of the sets of the form $f_\bN^{-1}(S)$ for measurable $S \subseteq X_J$. Hence the assumption implies that there is measurable $S \subseteq X_J$ such that $T = f_\bN^{-1}(S)$. Then we apply \Cref{thm:kolmog} with $s : X_J \to \{0,1\}$ the indicator function of $S$, which is a measurable map and therefore can also be regarded as a morphism in $\BorelStoch$, and $p = f_\bN P$. We therefore obtain that $sp : I \to \{0,1\}$ is deterministic. But since $sp$ corresponds to the probability measure on $\{0,1\}$ with weight $P(f_\bN^{-1}(S)) = P(T)$ on $\{1\}$ and the complementary weight $1 - P(T)$ on $0$, we conclude that indeed $P(T) \in \{0,1\}$.
\end{proof}

@@

\begin{corollary}[Hewitt-savage zero--one law]
    Suppose $\Omega$ is a standard Borel space with probability measure $P$, and let $\left(f_j: \Omega \to X\right)_{j\in \bN}$ be a sequence of independent, identically distributed random variables taking values in a standard Borel space $X$.
    Suppose $S \subset \prod_{j \in \bN} X = X_\bN$ is measurable in the product $\sigma$-algebra, and suppose moreover that, for each finite permutation $\alpha: \bN \to \bN$, $\hat{\alpha}(S) = S$ (in other words, $S$ is invariant under finite permutations of the variables).
    Then $P(f_\bN\inv(S)) \in \{0,1\}$ - here $f_\bN$ again denotes the induced map $\Omega \to X_\bN.$
\end{corollary}
\begin{proof}
    The proof is similar to the previous proof. This time we want to apply \Cref{thm:hewsav} with $\cC = \BorelStoch$, $J = \bN$, $A = I$.
    Again, we have $P: I \to \Omega$.
    As before, we let $s: X_\bN \to \{0,1\}$ be the indicator function of $S$, and we let $p = f_\bN P$.
    The i.i.d assumption on the $f_j$ implies that $\hat{\alpha}p = p$, and the assumption on $S$ implies $s\hat{\alpha} = s$ for finite permutations $\alpha$.
    Hence we can apply \Cref{thm:hewsav}. This again implies that the map $sp: I \to \{0,1\}$ is deterministic, which, just as before, means 
    $P(f_\bN\inv(S))$ is either $0$ or $1$.
\end{proof}
